TrenchR: an R package for transparent microclimate and biophysical ecology

### 2022-06-28 Manuscript_full.Rmd
Clean up code chunks

---
title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
author: 
  - "Lauren B. Buckley, Bryan A. Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy,"
  - "Abigail V. Meyer, Eric A. Riddell, Yutaro Sakairi, and Juniper Simonis"
output:
  html_document: default
  word_document: default
  pdf_document: 
    pandoc_args: --listings
    includes:
      in_header: preamble.tex
bibliography: bibliography.bib
csl: methods-in-ecology-and-evolution.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

doy   <- day_of_year("2021-06-01", format = "%Y-%m-%d") # Julian Date (DOY)

@@

# Separate solar radiation into direct, diffuse, and reflected

Sdir <- Srad[1,] # Direct solar radiation
Sdif <- Srad[2,] # Diffuse solar radiation
Sref <- Srad[3,] # Reflected solar radiation
```

@@

# Soil temperature (C)

@@

At the second stage, we use microclimate models to scale air temperature and wind speed from weather station height (2 m) to lizard height (0.02 m). 
We assume a surface roughness of $z_0 = 0.2$ m, which corresponds to bare sand and determines the turbulence of airflow. 
We implement a neutral air temperature profile driven by density differences, an air temperature profile forced by wind speed, and a neutral wind speed profile.

```{r}
# Scale air temperature to lizard height
# T_r = reference temperature
# zr  = reference height
# z0  = surface roughness
# z   = organism height
# T_s = surface temperature

Ta_liz  <- air_temp_profile_neutral(T_r = Ta, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts)  # Neutral air temperature profile
Ta_liz2 <- air_temp_profile(T_r = Ta, u_r = V, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts) # Air temperature profile forced by wind speed

# Scale wind speed to lizard height
# u_r = reference wind speed
# zr  = reference height
# z0  = surface roughness
# z   = organism height

V_liz <- wind_speed_profile_neutral(u_r = V, zr = 2, z0 = 0.2 , z = 0.02) # Neutral wind speed profile
```

@@

where $Q_{net}$ is the net energy exchange with the environment (W), $Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to convection (W), $Q_{cond}$ is energy exchange due to conduction (W), $Q_{met}$ is the energy emitted due to metabolism (W), and $Q_{evap}$ is the energy emitted due to evaporative water loss (W). 
We will estimate each term on the right side of the equation in turn. 
Estimating $Q_{abs}$ requires the surface area exposed to radiation and the solar absorptivity of the animal surface ($a$ proportion). 
We model a 10 gram *Sceloporus* lizard with solar absorptivity $a = 0.9$ [@gates1980biophysical]. 
We will initially assume $T_b = T_a$ to illustrate the calculations before solving for $T_b$ given the environmental conditions.

@@

We estimate thermal radiation $Q_{emit}$ (W) for the lizard outdoors. 
We assume the surface emissivity of lizards, $epsilon_s = 0.965$ [@barlett1967].

@@

  # psa_dir = proportions of surface area exposed to direct solar radiation
  # psi_ref = proportions of surface area exposed to reflected solar radiation
  # A       = surface area (m^2)

@@

We next estimate convection $Q_{conv}$ (W) and conduction $Q_{cond}$ (W). 
We will estimate the lizard's heat transfer coefficient, $H_L$ ($Wm^{-2}K^{-1}$) using an empirical relationship for lizards (`heat_transfer_coefficient()`). 
We also illustrate a function estimating $H_L$ using a spherical approximation (`heat_transfer_coefficient_approximation()`) and a simplified approximation (`heat_transfer_coefficient_simple()`) for cases when taxon specific relationships for estimating heat transfer coefficients are not available.

```{r}
ap <- airpressure_from_elev(elev) * 1000 # Barometric pressure (pascal)

# Use DRYAIR from NicheMapR to estimate the thermal conductivity of air and kinematic viscosity

@@

K  <- DRYAIRout$thcond # Thermal conductivity (Wm^-2K^-1) 
nu <- DRYAIRout$viskin # Kinematic viscosity (m2 s-1)

@@

# We will use the average of K and nu across the day for simplicity and since there is not a substantial diurnal variation.

K  <- mean(K)  # Average thermal conductivity across the day
nu <- mean(nu) # Average kinematic viscosity

@@

where an enhancement factor, $ef$, multiplier can be incorporated to account for increases in heat exchange resulting from air turbulence in field conditions. 
We implement the function in R assuming that 2/3 of the lizard's surface area is exchanging heat through convection. 
We currently estimate no convection since we are assuming $Tb = Ta$ for illustration purposes:

@@

We additionally estimate $T_b$ using a specialized function for lizards [@buckley2008link]:

@@

  # F_d, F_r, F_a, and F_g are the view factors between the surface of the lizard and diffuse solar radiation, reflected solar radiation, and atmospheric thermal radiation, and ground thermal radiation, respectively.

>>>>

---
title: 'TrenchR: an R package for transparent microclimate and biophysical ecology'
author: 
  - "Lauren B. Buckley, Bryan A. Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy,"
  - "Abigail V. Meyer, Eric A. Riddell, Yutaro Sakairi, and Juniper Simonis"
output:
  html_document: default
  word_document: default
  pdf_document: 
    pandoc_args: --listings
    includes:
      in_header: preamble.tex
bibliography: bibliography.bib
csl: methods-in-ecology-and-evolution.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

doy   <- day_of_year("2021-06-01", format = "%Y-%m-%d") # Day of year

@@

# Separate solar radiation into direct, diffuse, and reflected components

Sdir <- Srad[1,] # Direct solar radiation (W/m2)
Sdif <- Srad[2,] # Diffuse solar radiation  (W/m2)
Sref <- Srad[3,] # Reflected solar radiation  (W/m2)
```

@@

# Soil surface temperature (C)

@@

At the second stage, we use microclimate models to scale air temperature ($T_r$) and wind speed ($u_r$) from weather station height (reference height $z_r$= 2 m) to lizard height (organism height $z$= 0.02 m). 
We assume a surface roughness of $z_0 = 0.2$ m, which corresponds to bare sand and determines the turbulence of airflow. 
We implement neutral air temperature and wind speed profiles driven by density differences but profiles forced by wind speed are also available.

```{r}
# Neutral air temperature profile

Ta_liz  <- air_temp_profile_neutral(T_r = Ta, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts)

# Neutral wind speed profile

V_liz <- wind_speed_profile_neutral(u_r = V, zr = 2, z0 = 0.2 , z = 0.02) 
```

@@

where $Q_{net}$ is the net energy exchange with the environment (W), $Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to convection (W), $Q_{cond}$ is energy exchange due to conduction (W), $Q_{met}$ is the energy emitted due to metabolism (W), and $Q_{evap}$ is the energy emitted due to evaporative water loss (W). 
We will estimate each term on the right side of the equation in turn. 
Estimating $Q_{abs}$ requires the surface area exposed to radiation and the solar absorptivity of the animal surface ($a$ proportion). We use zenith angle $psi$ to estimate the projected (silhouette) area as a portion of the surface area of the organism, which allows estimating absorbed solar radiation.
We model a 10 gram *Sceloporus* lizard with solar absorptivity $a = 0.9$ [@gates1980biophysical]. 
We will initially assume $T_b = T_a$ to illustrate the calculations before solving for $T_b$ given the environmental conditions.

@@

We estimate thermal radiation $Q_{emit}$ (W) for the lizard outdoors, where $psa_{dir}$ and $psa_{ref}$ are the proportions of surface area $A$ ($m^2$) exposed to direct and reflected solar radiation, respectively. We assume the surface emissivity of lizards, $epsilon_s = 0.965$ [@barlett1967].

@@

@@

We next estimate convection $Q_{conv}$ (W) and conduction $Q_{cond}$ (W). 
We will estimate the lizard's heat transfer coefficient, $H_L$ ($Wm^{-2}K^{-1}$) using an empirical relationship for lizards (`heat_transfer_coefficient()`). We average thermal conductivity and kinematic viscosity across the day for simplicity and since there is not substantial diurnal variation. We also illustrate a function estimating $H_L$ using a spherical approximation (`heat_transfer_coefficient_approximation()`) and a simplified approximation (`heat_transfer_coefficient_simple()`) for cases when taxon specific relationships for estimating heat transfer coefficients are not available.

```{r}
# Use DRYAIR from NicheMapR to estimate the thermal conductivity of air and kinematic viscosity

ap <- airpressure_from_elev(elev) * 1000 # Barometric pressure (pascal)

@@

K  <- mean(DRYAIRout$thcond) # Thermal conductivity (Wm^-2K^-1) 
nu <- mean(DRYAIRout$viskin) # Kinematic viscosity (m2 s-1)

@@

@@

where an enhancement factor, $ef$, multiplier can be incorporated to account for increases in heat exchange resulting from air turbulence in field conditions. 
We implement the function in R assuming that 2/3 of the lizard's surface area is exchanging heat through convection. 
We currently estimate no convection since we are assuming $T_b = T_a$ for illustration purposes:

@@

We additionally estimate $T_b$ using a specialized function for lizards [@buckley2008link], where $F_d$, $F_r$, $F_a$, and $F_g$ are the view factors between the surface of the lizard and diffuse solar radiation, reflected solar radiation, atmospheric thermal radiation, and ground thermal radiation, respectively:

@@
